Writing this thesis and conducting the study has been nothing short of an adventure, and for this ride, I would like to first and foremost show my deepest gratitude to my advisor, Prof. Nimmi Rangaswamy. Her constant support as an advisor and friend in academic and personal matters has been invaluable. I would also like to thank her for introducing me to the field of HCI and Fairness in Social Applications of AI.

I have had the pleasure of working alongside several amazing people on this project, but I would like to thank my co-authors for my first publication, Pratham Gupta and Ben Varghese, without whose contribution this thesis would not have been possible. 

Next, I would like to thank my friends - Aakash Aanegola, Aakash Jain, Arihanth Tadanki, Dhruv Kapur, Kunwar Shaanjeet, Jammu(Aditya), Raghav Dwivedi, and Sidharth Giri, for the constant support you have provided me during my time at IIIT. They have been there for me when I have needed them the most, and I am eternally grateful to them. I want to thank Autrio, Aryan, Arnav, Bala, Haran, Himanshu, Hardik and Pavan for the countless hours of fun that have helped me relax and unwind. Pavani, thank you for being the fun gym partner you are. Sankeerthana, thank you for your fantastic company and those talks till sunrise. Shristy, Anushree, Ketan, Anamaya, Dhruv Verma, and Priyansh, thank you for your support straight from home. My CHD batchmates, for the fantastic memories we have made together. Also special mention to Chicken n Rice and ADHD.

Without all these people, my thesis wouldn't have been possible, and I'll forever be grateful to you.

Finally, I would like to thank my family, Mom, Dad, Aishni, and Arshia; I am who I am because of you. You have been my constant support, and I am eternally grateful to you for it. Thank you for believing in me.